### 2016-02-12 doc/thesis/methods.tex
Updated reading and writing

\subsection{Programs}

Most of the computer programs used for this work were written in the C
programming language. Contact networks were generated using the
\software{igraph} library~\autocite{csardi2006igraph}, which was also used to
store phylogenetic trees. Hash tables, and the dynamic programming matrices for
computing the tree kernel, were stored using Judy
arrays~\autocite{baskins2004judy}. The GNU Scientific Library
(GSL)~\autocite{gough2009gnu} was used to generate random draws from
probability distributions, and to perform the bisection step in the ABC-SMC
algorithm.

\subsubsection{Epidemic simulation over a network}

I implemented method of~\autocite{gillespie1977exact} for simulating epidemics,
and the corresponding transmission trees, over static contact networks. This
method has been applied previously for the same
purpose~\autocite{robinson2013dynamics, leventhal2012inferring}, however no
existing open source implementations were available.

Let $G = (V, E)$ be a directed contact network. Each directed edge $e = (u, v)$
in the network is associated with a transmission rate $\beta_e$, which indicates
that, once $u$ becomes infected, $u$ will infect $v$ with rate $\beta_e$. In
other words, the time between $u$ becoming infected and the infection passing
from $u$ to $v$ is a random variable with distribution $\Exponential(\beta_e)$.
Note that $v$ may become infected before this time has elapsed, if $v$ has
other incoming edges. Each vertex $v$ also has a removal rate $\gamma_v$, which
is the rate at which $v$ is removed from the population after becoming
infected. Removal may correspond to death or recovery with immunity, or a
combination of both, but in our implementation recovered nodes never re-enter
the susceptible population. Borrowing a term from HIV epidemiology, a
\defn{discordant edge} in $G$ is an edge $(u, v)$ where $u$ is infected and $v$
is uninfected.

To describe the algorithm, we introduce some notation and variables. Let
$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes.
Let $D$ be the set of discordant edges in the network, $\beta$ be the 
total transmission rate over all discordant edges, and $\gamma$ be the total
removal rate of all infected nodes. That is,
\[
  \beta = \sum_{e \in D} \beta_e, \quad
  \gamma = \sum_{v \in I} \gamma_v.
\]
The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
simulation progresses, but since these updates are very straightforward we do
not write them explicitly in the algorithm. When a new node $v$ is infected, it
is deleted from $S$ and added to $I$, any formerly discordant edges in $\in(v)$
are deleted from $D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$.
If $v$ is later removed, it is deleted from $I$ and added to $R$, and any
discordant edges in $\out(v)$ are deleted from $D$. In both cases, the
variables $\beta$ and $\gamma$ are updated to reflect the changes.

@@

The full algorithm is shown in Algorithm \ref{alg:nettree}. The transmission
tree $T$ is simulated along with the epidemic. We keep a map called $\tip$,
which maps infected nodes in $I$ to the tips of $T$. The variable $n$ counts
the number of nodes, both extant and internal, in $T$. The simulation continues
until either there are no discordant edges left in the network, or we reach a
user-defined cutoff of time ($t_{\max}$) or number of infections ($I_{\max}$).

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect one node $v$ at random
    \State let $T$ be a single node with index $1$
    \State $\tip[v] \gets 1$
    \State $n \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, 1) < \nicefrac{\beta}{\beta + \gamma}$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State add nodes with labels $n+1$ and $n+2$ to $T$
          \State connect $n+1$ and $n+2$ to $\tip[v]$ in $T$, with branch lengths $0$
          \State $\tip[v] \gets n + 1$
          \State $\tip[u] \gets n + 2$
          \State $n \gets n + 2$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State remove $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

\subsubsection{Adaptive sequential Monte-Carlo approximate Bayseian computation}

I implemented the adaptive sequential Monte-Carlo (SMC) algorithm developed by
Del Moral \etal~\autocite{del2012adaptive} for approximate Bayesian
computation.

For ease of exposition, we simplifiy the notation of Del Moral \etal by
dropping the subscripts on the variables which indicate the current iteration
of the algorithm. Instead, we will add a superscript $^*$ to indicate the value
of the variable which will be used in the next iteration (this should become
clear later on). Let $\set{\theta_i}_{i=1}^{N}$ be a set of $N$ particles,
which are particular choices of the model parameters $\theta$. For each
particle $\theta_i$, let $\set{X_{i,j}}_{j=1}^M$ be a set of $M$ simulated
datasets, generated from the model with parameters $\theta_i$. Let
$\set{W_i}_{i=1}^N$ be weights associated to each particle. Let $d$ be a
distance measure on data sets, so that $d(x, y)$ is smaller the more similar
$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
indicates whether a data set is ``close'' to the observed data. That is, if
$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
are distant.

>>>>

\subsection{Computer program}

I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. The program was written in the
\software{C} programming language. The \software{igraph}
library~\autocite{csardi2006igraph} was used to generate and store contact
networks and phylogenies. Judy arrays~\autocite{baskins2004judy} were used for
hash tables and dynamic programming arrays. The
\gls{GSL}~\autocite{gough2009gnu} was used to generate random draws from
probability distributions, and to perform the bisection step in the
\gls{ABC}-\gls{SMC} algorithm.

For ease of exposition, we simplifiy the notation of \textcite{del2012adaptive}
by dropping the subscripts on the variables which indicate the current
iteration number. Instead, we will add a prime $'$ to indicate a value which
will be used in the next iteration (this should become clear later on).

In the algorithm, we keep track of a population of $n$ sets of model
parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.

The particles are initially drawn from the prior distribution.

Let $d$ be a
distance measure on data sets, so that $d(x, y)$ is smaller the more similar
$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
indicates whether a data set is ``close'' to the observed data. That is, if
$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
are distant.

I implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics, and the corresponding transmission trees, over static
contact networks. This method has been independently implemented and applied by
several authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian}.
\textcite{groendyke2011bayesian} published their implementation as an
\software{R} package, but since the \gls{SMC} algorithm is quite
computationally intensive, we chose to implement our own version in
\software{C}.

Let $G = (V, E)$ be a directed contact network. The individual nodes and edges
of $G$ follow the dynamics of the \gls{SIR}
model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
the network is associated with a transmission rate $\beta_e$, which indicates
that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
distributed as $\Exponential(\beta_e)$. Note that $v$ may become infected
before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
removal rate $\gamma_v$, so that the waiting time until removal of $v$ from the
population is $\Exponential(\gamma_v)$. Removal may correspond to death or
recovery with immunity, or a combination of both, but in our implementation
recovered nodes never re-enter the susceptible population. We define a
\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$
has never been infected.

To describe the algorithm, we introduce some notation and variables. Let
$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes,
and $D$ be the set of discordant edges in the network. Let $\beta$ be the total
transmission rate over all discordant edges, and $\gamma$ be the total removal
rate of all infected nodes,
\[
  \beta = \sum_{e \in D} \beta_e, \quad
  \gamma = \sum_{v \in I} \gamma_v.
\]
The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
and added to $I$, any formerly discordant edges in $\in(v)$ are deleted from
$D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
removed, it is deleted from $I$ and added to $R$, and any discordant edges in
$\out(v)$ are deleted from $D$. In both cases, the variables $\beta$ and
$\gamma$ are updated to reflect the changes. Since these updates are
straightforward, we do not write them explicitly in the algorithm.

@@

The Gillespie simulation algorithm is given as Algorithm~\ref{alg:nettree}. The
transmission tree $T$ is simulated along with the epidemic. We keep a map
called $\tip$, which maps infected nodes in $I$ to the tips of $T$. The
simulation continues until either there are no discordant edges left in the
network, or we reach a user-defined cutoff of time ($t_{\max}$) or number of
infections ($I_{\max}$). We use the notation $\Uniform(0, 1)$ to indicate a
number drawn from a uniform distribution on $(0, 1)$, and likewise for
$\Exponential(\lambda)$. The combined number of internal nodes and tips in $T$
is denoted $|T|$.

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
    \State $T \gets$ a single node with label $1$
    \State $\tip[v] \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, \beta + \gamma) < \beta$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
          \Comment $|T|$ increased by 2
          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
          \State $\tip[v] \gets |T|-1$
          \State $\tip[u] \gets |T|$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State delete $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@
